---
title: "AKQoL Item Fit Analysis"
subtitle: "Empirical critical values for polytomous Rasch item fit statistics"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document:
    latex_engine: pdflatex
    number_sections: true
geometry: margin=2.5cm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center")
library(RASCHplot)
library(ggplot2)
library(knitr)
library(ggpubr)
load("ak_final_analysis.RData")
theme_set(theme_minimal() + theme(legend.title = element_blank(),
                                   plot.title = element_text(size = 9, hjust = 0.5),
                                   text = element_text(size = 8)))

# RASCHplot's own plot.RASCHstats() calls density() on the raw min/max
# values without filtering non-finite ones. Outfit/Infit/FitResid can be
# +/-Inf when an item's total weighted variance is (numerically) zero in a
# given replicate, so this rebuilds the same quantile-shaded density plot
# but drops non-finite values first, and renders a placeholder instead of
# crashing when a distribution is degenerate (too few distinct finite
# values to estimate a density from).
safe_null_plot <- function(sim, type, extreme, probs, title) {
  which_item <- sapply(sim$statobj, function(s) items[if (extreme == "min") which.min(s[[type]]) else which.max(s[[type]])])
  z <- sapply(sim$statobj, function(s) if (extreme == "min") min(s[[type]], na.rm = TRUE) else max(s[[type]], na.rm = TRUE))
  n_total <- length(z)
  finite_z <- z[is.finite(z)]
  n_dropped <- n_total - length(finite_z)

  # Degenerate: fewer than 2 distinct finite values means there is no
  # distribution to draw a density for (either every replicate is
  # non-finite, e.g. Infit = Inf, or every replicate lands on the exact
  # same finite value, e.g. Outfit = 0 whenever some item's simulated data
  # has zero weighted residual variance).
  if (length(unique(finite_z)) < 2) {
    driver <- sort(table(which_item), decreasing = TRUE)
    driver_txt <- paste(names(driver), " (", driver, "/", n_total, ")", sep = "", collapse = ", ")
    val_txt <- if (length(finite_z) == 0) "non-finite" else paste0("= ", signif(unique(finite_z), 3))
    msg <- paste(c(paste0(extreme, " ", type, " is ", val_txt), paste0("in all ", n_total, " replicates."),
                   "Item achieving it:", strwrap(driver_txt, width = 24)), collapse = "\n")
    return(ggplot() + annotate("text", x = 0, y = 0, size = 2.3, label = msg) +
      xlim(-1, 1) + ylim(-1, 1) + theme_void() + ggtitle(title))
  }

  # A handful of replicates can inflate to astronomically large values (a
  # near-singular refit on an already-fragile item), so a heavy-tailed
  # distribution is plotted on a log10 x-axis - otherwise the density is
  # compressed into an unreadable sliver near zero.
  use_log <- all(finite_z > 0) && (max(finite_z) / min(finite_z) > 1000)
  z_plot <- if (use_log) log10(finite_z) else finite_z
  dens <- density(z_plot)
  qs <- quantile(z_plot, probs = probs)
  breaks <- if (extreme == "min") paste0(probs * 100, "%") else paste0((1 - probs) * 100, "%")
  labels <- if (extreme == "min") c(breaks, "other") else c("other", breaks)
  colours <- if (extreme == "min") c("#A43340", "#F44F4F", "#CFCFC2") else c("#CFCFC2", "#F44F4F", "#A43340")
  names(colours) <- labels
  ints <- findInterval(dens$x, qs)
  df <- data.frame(x = dens$x, y = dens$y, quant = factor(labels[ints + 1], levels = labels))
  xt <- paste0(ifelse(extreme == "max", "Maximal ", "Minimal "), type, if (use_log) " (log10)" else "")
  cap <- if (n_dropped > 0) paste0(title, " (", n_dropped, "/", n_total, " excl.)") else title
  ggplot(df, aes(x, y)) +
    geom_area(aes(fill = quant), position = "identity") +
    scale_fill_manual(values = colours, name = "Quantiles") +
    xlab(xt) + ylab("") + ggtitle(cap) +
    theme(axis.ticks.y = element_blank(), axis.text.y = element_blank())
}
```

# Data and method

The AKQoL (actinic keratosis quality of life) questionnaire consists of 10 items
(`Q1`-`Q10`), originally scored 0-3. Following a check of category usage,
categories 2 and 3 were collapsed for all 10 items, giving a 0-2 scale.
`r length(extreme_idx)` respondent(s) with an extreme (minimum or maximum
possible) total score were excluded, since their person-parameter estimate is
undefined, leaving **N = `r nrow(dat)`** respondents.

Item and person parameters were estimated by conditional maximum likelihood
(CML, via `eRm::PCM`) and weighted maximum likelihood (WML) respectively,
using the `RASCHplot` package (v`r as.character(packageVersion("RASCHplot"))`)
with the local patch (`RASCHplot_patches.R`) applied — this fixes an item/
threshold orientation bug in `pcmfct()`/`rRMPstats()` and a numerical-overflow
issue that otherwise crashes the simulation for polytomous data. The empirical
sampling distribution of the minimum and maximum item fit statistics across
all `r ncol(dat)` items was estimated from **B = `r B`** parametric bootstrap
replicates (`rRMPstats()`), refitting CML/WML on each simulated data set.

# Item parameters

```{r item-params}
delta_df <- data.frame(Item = items, round(delta, 3))
colnames(delta_df) <- c("Item", paste0("Threshold ", 1:ncol(delta)))
kable(delta_df, row.names = FALSE,
      caption = "Estimated item step (threshold) parameters (CML).")
```

# Evaluation of item fit

Observed fit statistics for each item, compared against the empirical null
distribution of the minimum/maximum across items from the `r B` simulated
replicates. `*` marks an item beyond the 5%/95% empirical percentile,
`**` beyond the 2.5%/97.5% percentile (two-sided, unadjusted for multiple
testing).

```{r fit-eval}
pctl <- function(type, extreme, probs) {
  z <- sapply(sim$statobj, function(s) if (extreme == "min") min(s[[type]], na.rm = TRUE) else max(s[[type]], na.rm = TRUE))
  z <- z[is.finite(z)]
  quantile(z, probs = probs, na.rm = TRUE)
}
flag <- function(x, type) {
  lo_25 <- pctl(type, "min", 0.025); lo_5 <- pctl(type, "min", 0.05)
  hi_95 <- pctl(type, "max", 0.95); hi_975 <- pctl(type, "max", 0.975)
  is_below <- function(v, thr) !is.na(thr) & v < thr
  is_above <- function(v, thr) !is.na(thr) & v > thr
  ifelse(is_below(x, lo_25) | is_above(x, hi_975), "**",
         ifelse(is_below(x, lo_5) | is_above(x, hi_95), "*", ""))
}

fit_df <- data.frame(
  Item = items,
  Outfit = sprintf("%.3f%s", obs$Outfit, flag(obs$Outfit, "Outfit")),
  Infit = sprintf("%.3f%s", obs$Infit, flag(obs$Infit, "Infit")),
  tOutfit = round(obs$tOutfit, 3),
  tInfit = round(obs$tInfit, 3),
  FitResid = round(obs$FitResid, 3)
)
kable(fit_df, row.names = FALSE,
      caption = "Observed item fit statistics, flagged against the empirical null. Outfit is not flagged against a maximum threshold, and Infit not against a minimum threshold, since those reference distributions are degenerate (see Section 4).")
```

# Simulated empirical null distributions

The minimal-Outfit and maximal-Infit distributions are both degenerate under
this model, for the same reason: in every one of the `r B` replicates, at
least one item has zero total weighted residual variance in the simulated
data set, giving that item Outfit $=0$ and Infit $=\infty$ simultaneously.
This is not a fixed item - it rotates among Q3, Q4, Q5, Q7, Q8 and Q9 (the
same items flagged in Table 2), confirming that their extreme observed fit
reflects a genuinely fragile CML calibration for this collapsed scale rather
than a one-off artifact. Both are shown as such below rather than as a
(misleading) density plot. The maximal-Outfit distribution is not degenerate,
but its right tail is extreme (up to the order of $10^{90}$ in a handful of
replicates, again traceable to the same fragile items being refit on
resampled data), so it is plotted on a log10 x-axis; even so, its median
(`r round(median(sapply(sim$statobj, function(s) max(s$Outfit))), 0)`) is far
above what an Outfit statistic should be under good fit, so this null
distribution should not be treated as a usable empirical critical value
without first addressing why Q3-Q9 refit so unstably.

```{r null-dist-outfit, fig.width=6.3, fig.height=2.6, fig.cap="Empirical null distribution of the minimum and maximum Outfit across items."}
p1 <- safe_null_plot(sim, "Outfit", "min", c(0.025, 0.05), "Minimal Outfit")
p2 <- safe_null_plot(sim, "Outfit", "max", c(0.95, 0.975), "Maximal Outfit")
ggarrange(p1, p2, ncol = 2)
```

```{r null-dist-infit, fig.width=6.3, fig.height=2.6, fig.cap="Empirical null distribution of the minimum and maximum Infit across items."}
p3 <- safe_null_plot(sim, "Infit", "min", c(0.025, 0.05), "Minimal Infit")
p4 <- safe_null_plot(sim, "Infit", "max", c(0.95, 0.975), "Maximal Infit")
ggarrange(p3, p4, ncol = 2)
```

```{r null-dist-fitresid, fig.width=6.3, fig.height=2.6, fig.cap="Empirical null distribution of the minimum and maximum FitResid across items."}
p5 <- safe_null_plot(sim, "FitResid", "min", c(0.025, 0.05), "Minimal FitResid")
p6 <- safe_null_plot(sim, "FitResid", "max", c(0.95, 0.975), "Maximal FitResid")
ggarrange(p5, p6, ncol = 2)
```
